---
title: "Analysis of Pokemon Stats"
subtitle: "For all pokemon of Generations I through VII"
author: "Natalie Weaver"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  xaringan::moon_reader:
    css: ["default", "metropolis", "metropolis-fonts", "my-css.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
---
exclude: true

```{r, setup, include = FALSE}
# load packages
library(pacman)
p_load(tidyverse, magrittr, janitor, DT, lubridate, viridis, knitr, cowplot)

# knitr options
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  dpi = 500)

options(knitr.table.format = "html")

```

---

## Introduction

Each pokemon has **stats** that determine its abilities in various categories:

* HP (hit points)
* Attack
* Defense

Each pokemon has a **primary type** based on its physiology:<sup>✨</sup>
.footnote[
✨ and just its general vibe
]

- Water
- Fire
- Rock

Which type of pokemon is the strongest, based on its stats?

---

## Packages and Data

```{r, echo = TRUE, eval = FALSE}
# load the packages we need for this analysis
library(pacman)
p_load(tidyverse, magrittr, viridis)
```

```{r}
# load, manipulate, and show the data
pokemon <- read_csv("pokemon.csv")
pok_by_type <- pokemon %>% group_by(type1) %>%
  summarize(
    mean_hp = mean(hp),
    mean_attack = mean(attack),
    mean_sp_attack = mean(sp_attack),
    mean_defense = mean(defense),
    mean_sp_defense = mean(sp_defense),
    mean_speed = mean(speed)
  )
pok_by_type %>% head() %>% 
  datatable(
    rownames = FALSE,
    colnames = c("Type", "HP", "Attack", "Sp. Attack", "Defense", "Sp. Defense", "Speed"),
    options = list(dom = 't', scrollX='400px')
  ) %>% 
  formatRound(columns = 2:7, digits = 3)
```

---

## Stats Breakdown by Type

```{r, fig.height = 5.75}
pok_by_type_longer <- pok_by_type %>% pivot_longer(
  cols = !type1,
  names_to = "stat",
  values_to = "value"
)

ggplot(data = pok_by_type_longer, aes(x = type1, y = value)) +
  geom_bar(stat = "identity", aes(fill = stat)) +
  guides(x = guide_axis(angle = 90)) +
  scale_fill_viridis(
    discrete = TRUE,
    labels = c("Attack", "Defense", "HP", "Special Attack", "Special Defense", "Speed")
  ) +
  labs(
    x = "Primary Type",
    y = "Mean of Total Statistics",
    fill = "Statistic"
  ) +
  theme(text = element_text(size = 20))

```

---

## Distribution of Stat Values

```{r}
# distribution of attack
attack_hist <- ggplot(data = pokemon, aes(x = attack)) +
  geom_histogram(bins = 20) +
  theme(text = element_text(size = 10))

# distribution of special attack
sp_attack_hist <- ggplot(data = pokemon, aes(x = sp_attack)) +
  geom_histogram(bins = 20) +
  theme(text = element_text(size = 10))

# distribution of defense
defense_hist <- ggplot(data = pokemon, aes(x = defense)) +
  geom_histogram(bins = 20) +
  theme(text = element_text(size = 10))

# distribution of special defense
sp_defense_hist <- ggplot(data = pokemon, aes(x = sp_defense)) +
  geom_histogram(bins = 20) +
  theme(text = element_text(size = 10))

# distribution of speed
speed_hist <- ggplot(data = pokemon, aes(x = speed)) +
  geom_histogram(bins = 20) +
  theme(text = element_text(size = 10))

# distribution of hp
hp_hist <- ggplot(data = pokemon, aes(x = hp)) +
  geom_histogram(bins = 20) +
  theme(text = element_text(size = 10))

plot_grid(attack_hist, sp_attack_hist, defense_hist, sp_defense_hist, speed_hist, hp_hist, NULL, NULL, ncol = 2) 
```